noam_data_exploration_v1

### 2024-07-17 noam_data_exploration_v3.Rmd
Looked at transect level diversity, stability, synchrony stuff. Redid model selection at transect level. Need to sort issues for work at quadrat level.

```{r transect level portfolio effect}
# using Anderson et al. 2013 definitions of portfolio effects - relevant fxns in ecofolio packages
wide_transect_level$Stratum <- as.factor(paste(wide_transect_level$Site, "-", wide_transect_level$Habitat, "-", wide_transect_level$Transect))
x <- subset(wide_transect_level, Stratum == "LTER 1 - Backreef - 1")

y <- x[ , -which(names(x) %in% meta_cols)]
y <- y[, which(colSums(y) != 0)]
z <- pe_avg_cv(y, ci = TRUE, detrending = "loess_detrended")

@@

diversity_stability <- merge(diversity_stability, lm_synchrony_transect, by = "Stratum") # add in loreau and de Mazancourt synchrony
diversity_stability <- merge(diversity_stability, gross_synchrony_transect, by = "Stratum")
diversity_stability <- merge(diversity_stability, transect_level_portfolio, by = "Stratum") # add in portfolio effects
pairs(~ Cover_mean + Cover_cv + Cover_stability + 
        Shannon_mean + Shannon_cv + Shannon_stability + 
        Loreau_synchrony + Gross_synchrony + Portfolio_effect, data = diversity_stability)

diversity_mod_1 <- glmmTMB(Cover_cv ~ Richness_mean, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_1)
plot(allEffects(diversity_mod_1)) # richness reduces cv (i.e. enhances stability!)

diversity_mod_2 <- glmmTMB(Cover_cv ~ Shannon_mean, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_2) # similar results

diversity_mod_3 <- glmmTMB(Cover_cv ~ Habitat*Site, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_3) # christ 
Anova(diversity_mod_3) # all highly sig but what the fuck am I doing here

diversity_mod_4 <- glmmTMB(Cover_cv ~ Habitat, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_4)
Anova(diversity_mod_4)
plot(allEffects(diversity_mod_4)) # greatest variance in the lagoon (makes sense given beta diversity trends above)

diversity_mod_5 <- glmmTMB(Cover_cv ~ Site, family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_5)
Anova(diversity_mod_5)

diversity_mod_6 <- glmmTMB(Cover_cv ~ Shannon_mean*Habitat + (1|Site), family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_6) # negligible variance in random effects
Anova(diversity_mod_6) # Diversity the strongest predictor of "stability" by far!

diversity_mod_7 <- glmmTMB(Cover_cv ~ Shannon_mean*Site + (1|Site_habitat), family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_mod_7)
Anova(diversity_mod_7)

diversity_mod_universal <- glmmTMB(Cover_cv ~ Shannon_mean*Site*Habitat + (1|Site_habitat), family = Gamma(link = "log"), data = diversity_stability, na.action = na.pass)
diversity_dredge <- dredge(diversity_mod_universal)
model.sel(diversity_dredge)

# best model based on dredge:
diversity_best <-  glmmTMB(Cover_cv ~ Shannon_mean*Site + Site*Habitat + (1|Site_habitat), family = Gamma(link = "log"), data = diversity_stability)
summary(diversity_best) # very low variance in random effect, which is great
Anova(diversity_best) # Interactions all significant
plot(dp <- simulateResiduals(diversity_best)) # some outliers, but otherwise good model fit. I am fine with this. 
r.squaredGLMM(diversity_best) # 0.77, 0.77

(shannon_cv_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Cover_cv, colour = Site, shape = Habitat), size = 8) + 
   scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
    stat_smooth(aes(x = Shannon_mean, y = Shannon_cv),method = "lm", se = T, size = 2) + 
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

(shannon_cv_plot_2 = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Cover_cv, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

@@

# GOALS: test portfolio effects, look at turnover and temporal beta diversity stats, ways to test beta dispersion

@@

@@

```{r diversity stability quadrat level}

>>>>

```{r transect level portfolio effect}
# using Anderson et al. 2013 definitions of portfolio effects - relevant fxns in ecofolio packages
wide_transect_level$Stratum <- as.factor(paste(wide_transect_level$Site, "-", wide_transect_level$Habitat, "-", wide_transect_level$Transect))

@@

diversity_stability <- merge(diversity_stability, lm_synchrony_transect, by = "Stratum") # add in loreau and de Mazancourt synchrony
diversity_stability <- merge(diversity_stability, gross_synchrony_transect, by = "Stratum")
diversity_stability <- merge(diversity_stability, transect_level_portfolio, by = "Stratum") # add in portfolio effects
pairs(~ Cover_mean + Cover_cv + Cover_stability + 
        Shannon_mean + Shannon_cv + Shannon_stability + 
        Loreau_synchrony + Gross_synchrony + Portfolio_effect, data = diversity_stability)
```

```{r basic diversity stability plots transect level, fig.height=8, fig.width=10}
(shannon_cv_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Cover_stability, colour = Site, shape = Habitat), size = 8) + 
   scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
#    stat_smooth(aes(x = Shannon_mean, y = Shannon_cv),method = "lm", se = T, size = 2) + 
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

(shannon_cv_plot_2 = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Cover_stability, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

@@

@@

```{r portfolio effects and synchrony transect level,  fig.height = 8, fig.width = 10}
(loreau_stability_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Loreau_synchrony, y = Cover_stability, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

(gross_stability_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Gross_synchrony, y = Cover_stability, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

(portfolio_stability_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Portfolio_effect, y = Cover_stability, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         ) # almost a perfect line, which makes sense given that CV is a component of portfolio effects
```

```{r diversity, portfolio effects and synchrony transect level,  fig.height = 8, fig.width = 10}
(loreau_diversity_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Loreau_synchrony, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

(gross_diversity_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Gross_synchrony, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

(portfolio_diversity_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Portfolio_effect, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )
```

```{r model selection}
diversity_mod_universal <- glmmTMB(Cover_stability ~ Loreau_synchrony + Shannon_mean*Site*Habitat, family = Gamma(link = "log"), data = diversity_stability, na.action = na.pass)
diversity_dredge <- dredge(diversity_mod_universal)
model.sel(diversity_dredge)

# top model is as follows (by a decent margin):
diversity_top <- glmmTMB(Cover_stability ~ Loreau_synchrony + Shannon_mean*Site + Site*Habitat, family = Gamma(link = "log"), data = diversity_stability, na.action = na.pass)
summary(diversity_top)
Anova(diversity_top)
```

@@

```{r diversity stability quadrat level setup}
data$Stratum <- as.factor(paste(data$Site, "-", data$Habitat, "-", data$Transect))
data$Location <- as.factor(data$Location)

quadrat_portfolio_columns <- c("Location", "Portfolio_effect") # names of columns for dataframe
quadrat_level_portfolio <- 
  data.frame(matrix(nrow = length(unique(data$Location)), 
                    ncol = length(quadrat_portfolio_columns))
             )
colnames(quadrat_level_portfolio) <- quadrat_portfolio_columns # change column 
quadrat_level_portfolio$Location <- unique(data$Location) # populate

for(i in 1:length(unique(unique(data$Location))) ){ # for each  quadrat
  # create a new dataframe for current quadrat
  newData <- subset(data, Location == levels(data$Location)[i])
  current_matrix <- newData[ , -which(names(newData) %in% meta_cols)] # remove metadata columns
  if(sum(current_matrix) == 0){ # if this was a 0 algae quadrat
    quadrat_level_portfolio[i, 2] <- NA # set portfolio effects to 0
  }else{ # otherwise
    current_matrix <- current_matrix[, which(colSums(current_matrix) != 0)] # remove irrelevant algae

    quadrat_level_portfolio[i, 2] <- pe_avg_cv(current_matrix) # calculate portfolio effects for current time series
  }
